---
title: "Random Forest Classification for School Indicator"
output: html_document
---

# Random Forest Classification for School Indicator

## Set Up

### Packages

```{r, quietly = TRUE}
library(readr)
library(terra)
library(randomForest)
library(datasets)
library(caret)
library(sf)
library(dplyr)
library(ranger)  
library(kableExtra)
library(caret)
library(rmarkdown)
```

### Load in Data

Africa grids 10km x 10km using GHSL grids and data

Set seed

```{r}

setwd("/home/azureuser/cloudfiles/code/Users/ariley/FINAL")

grid_data <- read.csv("Data/Africa_Grid_Data_10km_ghsl.csv")

set.seed(23)

```

## Individual African Country Models

### Set-Up Data

Load in data, remove any missing data, name the variables something slightly more sensible, specify countries, and create factor indicator for schools/no schools.

```{r}

data <- grid_data[,-1]
data <- na.omit(data)
head(data)
names(data)[c(9:16)] <- c("smod.10", "smod.11", "smod.12", "smod.13", "smod.21", "smod.22", "smod.23", "smod.30")

countries <- c("BEN", "BWA", "GHA", "GIN", "KEN", "MWI", "NAM", "NER", "NGA", "RWA", "SLE", "SSD", "ZAF", "ZWE")

data$SCHOOLS <- as.factor(ifelse(data$NUMPOINTS > 0, 1, 0))

```

## Fit Initial Random Forest Classification Models

1. Subset individual country data
2. Sample testing and training set (training fraction = 0.7)
3. Run Random Forest model with formula
  $$
  SCHOOLS ~ x + y + land + pop +  built_s + built_v + smod.10 + smod.11 + smod.12 + smod.13 + smod.21 + smod.22 + smod.23 + smod.30
  $$
4. Save to file

```{r}

country.data <- list()
ind <- list()
train <- list()
test <- list()
rf <- list()
for (i in 1:length(countries)) {
    country.data[[i]] <- subset(data, country == i)
    ind[[i]] <- sample(2, nrow(country.data[[i]]), replace = TRUE, prob = c(0.7, 0.3))
    train[[i]] <- country.data[[i]][ind[[i]]==1,]
    test[[i]] <- country.data[[i]][ind[[i]]==2,]

    rf[[i]] <- ranger(SCHOOLS ~ x + y + land + pop + built_s + built_v + smod.10 + smod.11 + smod.12 + smod.13 + smod.21 + smod.22 + smod.23 + smod.30, 
                data = train[[i]], 
                respect.unordered.factors = "order",
                seed = 23)


    saveRDS(rf[[i]], file = paste("Outputs/", countries[[i]], "_10km_school_count_RF_AllVars_ghsl.rds", sep = ""))
    #rf[[i]] <- readRDS(file = paste("Outputs/", countries[[i]], "_10km_school_count_RF_Default.rds", sep = ""))

}

```

## Model Tuning

For a Random Forest classification model using the RandomForest package, the default hyperparameter variables are:
- $n_features = number of model covariates$
- $mtry = floor(sqrt(n_features))$
- $replace = TRUE$
- $min.node.size = 1$
- $sample.fraction$ from data set-Up
- $ntree = 500$

We consider a grid of alternative hyperparamters:
- $n_features = number of model covariates$
- $mtry = c(seq(1, n_features, 1))$
- $replace = c(TRUE, FALSE)$
- $min.node.size = c(seq(1, n_features, 1))$
- $sample.fraction = c(0.5, 0.6, 0.7, 0.8, 0.9)$
- $ntree = c(100, 200, 500, 1000)$

```{r}

n_features <- 14

hyper_grid <- expand.grid(
  mtry = c(seq(1, n_features, 1)), 
  min.node.size = c(seq(1, n_features, 1)),
  replace = c(TRUE, FALSE),
  sample.fraction = c(0.5, 0.6, 0.7, 0.8, 0.9),
  num.trees = c(50, 100, 200, 500),
  rmse = NA
)

```

Using a full grid search for each country, we run the model tuning. 
We aim to find the hyperparameter values that minimise the model RMSE.

```{r}

tuned_hyper <- list()
hyper_grid_list <- list()
for (i in 1:length(countries)){
    hyper_grid_list[[i]] <- hyper_grid
    
    set.seed(23)
    for(j in seq_len(nrow(hyper_grid_list[[i]]))){
        fit <- ranger(SCHOOLS ~ x + y + land + pop + built_s + built_v + smod.10 + smod.11 + smod.12 + smod.13 + smod.21 + smod.22 + smod.23 + smod.30, 
                    data = country.data[[i]],
                    num.trees = hyper_grid_list[[i]]$num.trees[j],
                    mtry = hyper_grid_list[[i]]$mtry[j],
                    min.node.size = hyper_grid_list[[i]]$min.node.size[j],
                    replace = hyper_grid_list[[i]]$replace[j],
                    sample.fraction = hyper_grid_list[[i]]$sample.fraction[j],
                    verbose = FALSE,
                    seed = 23,
                    respect.unordered.factors = 'order'
                    )
            hyper_grid_list[[i]]$rmse[j] <- sqrt(fit$prediction.error)
        }

    default_rmse <- sqrt(rf[[i]]$prediction.error)

    tuned_hyper[[i]] <- hyper_grid_list[[i]] %>% arrange(rmse) %>%
        mutate(perc_gain = (default_rmse - rmse) / default_rmse * 100) %>%
        head(1)
    
    tuned_hyper[[i]]$default_rmse <- sqrt(rf[[i]]$prediction.error)
    
    print(tuned_hyper[[i]])
    saveRDS(rf2[[i]], file = paste("Outputs/", countries[[i]], "_10km_school_indicator_RF_Tuned.rds", sep = ""))


}


```

We report the tuned hyperparameters, the related RMSE and the percentage gain in RMSE

```{r}

tuned_hyper %>%
  kbl() %>%
  kable_styling()

```

Now 

```{r}

rf2 <- list()

for (i in 1:length(country)){
    rf2[[i]] <- ranger(SCHOOLS ~ x + y + land + pop + built_s + built_v + smod.10 + smod.11 + smod.12 + smod.13 + smod.21 + smod.22 + smod.23 + smod.30, 
        data = train[[i]],
        mtry = tuned_hyper[[i]][1,1], 
        min.node.size = tuned_hype[[i]]r[1,2], 
        replace = tuned_hyper[[i]][1,3], 
        num.trees = tuned_hyper[[i]][1,5],
        respect.unordered.factors = "order",
        seed = 23
        )
    tuned_hyper[[i]]$rmse.final <- sqrt(rf2[[i]]$prediction.error)
    saveRDS(rf2[[i]], file = paste()"Outputs/", countries[[i]], "_10km_school_indicator_RF_Tuned.rds", sep = "")
}

```


## Prediction 

Using the model prediction function, we can predict 4 different values:
1. Prediction on training set
2. Prediction on testing set
3. Prediction on entire country data

We can then view the confusion matrix and model statistics, including:

- Confusion Matrix
- Accuracy
- 95\% CI
- No info rate
- p-value (acc > nir)
- kappa
- Mcnemar's Test P-Value
- sensitivity
- Specificity
- pos pred value
- neg pred value
- prevalence
- detection rate
- detection prevalence
- balanced accuracy
- positive class

```{r}

Africa <- st_read(dsn = "Data/Africa_Shapefile/Africa_Boundaries.shp")
st_crs(Africa)
#Africa <- st_transform(Africa, crs = "ESRI:54009")

head(Africa)

p1 <- list()
trained <- list()
p2 <- list()
tested <- list()
p3 <- list()
predicted <- list()
#AOI <- list()
Africa_54009 <- st_transform(Africa, crs = "ESRI:54009")
AOI <- subset(Africa_54009, ISO %in% countries)

pred.data <- list()
predicted.r <- list()
pred.sf <- list()

for (i in 1:length(countries)){
    p1[[i]] <- predict(rf2[[i]], train[[i]])
    trained[[i]] <- cbind(p1[[i]]$predictions, train[[i]])
    names(trained[[i]])[1] <- "p1"
    confusionMatrix(trained[[i]]$p1, train[[i]]$SCHOOLS)

    p2[[i]] <- predict(rf2[[i]], test[[i]])
    tested[[i]] <- cbind(p2[[i]]$predictions, test[[i]])
    names(tested[[i]])[1] <- "p2"
    confusionMatrix(tested[[i]]$p2, test[[i]]$SCHOOLS)

    p3[[i]] <- predict(rf2[[i]], country.data[[i]])
    predicted[[i]] <- cbind(p3[[i]]$predictions, country.data[[i]])
    names(predicted[[i]])[1] <- "p3"
    confusionMatrix(predicted[[i]]$p3, country.data[[i]]$SCHOOLS)
    predicted[[i]]$diff <- as.factor(as.numeric(predicted[[i]]$p3) - as.numeric(predicted[[i]]$SCHOOLS))


    #AOI[[i]] <- subset(Africa, ISO3 == countries[[i]])
    #AOI[[i]] <- st_transform(AOI[[i]], crs = "ESRI:54009")

    pred.sf[[i]] <- st_as_sf(predicted[[i]], coords = c("x","y"), crs = "ESRI:54009")
    pred.country <- st_intersection(pred.sf[[i]], AOI)

    pred.data[[i]] <- cbind(st_drop_geometry(pred.country), st_coordinates(pred.country))
}

pred.all <- do.call(rbind, pred.data)

saveRDS(pred.all, file = "Outputs/Africa_10km_school_count_RF_Pred.RDS")

```

From prediction 3., we can create the gridded prediction as a raster

```{r}

names(pred.data[[1]])
predicted.r <- list()
for (i in 1:length(countries)){
    data <- pred.data[[i]][,c(25,26,1:24)]
    predicted.r[[i]] <- rast(data, type = "xyz", crs = "ESRI:54009")
    country.r <- predicted.r[[i]]
    saveRDS(country.r, file = paste("Outputs/", countries[[i]], "_10km_school_count_RF_Pred_Raster.RDS", sep = ""))

}
plot(predicted.r[[1]])

```

We report the confusion matrices, accuracy, sensitivity, and specificity

```{r}
head(confusionMatrix(trained[[i]]$p1, train[[i]]$SCHOOLS))

CM <- list()
S <- list()
for (i in 1:length(countries)){
    CM[[i]] <- list()
    S[[i]] <- list()

    CM[[i]][[1]] <- confusionMatrix(trained[[i]]$p1, train[[i]]$SCHOOLS)$table
    S[[i]][[1]] <- confusionMatrix(trained[[i]]$p1, train[[i]]$SCHOOLS)$overall[1]
    S[[i]][[2]] <- confusionMatrix(trained[[i]]$p1, train[[i]]$SCHOOLS)$byClass[1]
    S[[i]][[3]] <- confusionMatrix(trained[[i]]$p1, train[[i]]$SCHOOLS)$byClass[2]

    CM[[i]][[2]] <- confusionMatrix(tested[[i]]$p2, test[[i]]$SCHOOLS)$table
    S[[i]][[4]] <- confusionMatrix(tested[[i]]$p2, test[[i]]$SCHOOLS)$overall[1]
    S[[i]][[5]] <- confusionMatrix(tested[[i]]$p2, test[[i]]$SCHOOLS)$byClass[1]
    S[[i]][[6]] <- confusionMatrix(tested[[i]]$p2, test[[i]]$SCHOOLS)$byClass[2]

    CM[[i]][[3]] <- confusionMatrix(predicted[[i]]$p3, country.data[[i]]$SCHOOLS)$table
    S[[i]][[7]] <- confusionMatrix(predicted[[i]]$p3, country.data[[i]]$SCHOOLS)$overall[1]
    S[[i]][[8]] <- confusionMatrix(predicted[[i]]$p3, country.data[[i]]$SCHOOLS)$byClass[1]
    S[[i]][[9]] <- confusionMatrix(predicted[[i]]$p3, country.data[[i]]$SCHOOLS)$byClass[2]

}

CM
S

```


And:

4. Probability prediction on entire country data

First we fit the Random Forest using `probability = TRUE` in the ranger function.
Then one loop we also create the gridded predictions as a raster and consider different thresholds of the predictive probabilities.
This is to create a priority order for other work. We consider significance at 6 different levels: 20%, 10%, 5%, 2.5%, 1%, 0.1%

```{r}



rf3 <- list()
for (i in 1:length(countries)) {
    rf3[[i]] <- ranger(SCHOOLS ~ x + y + land + pop + built_s + built_v + smod.10 + smod.11 + smod.12 + smod.13 + smod.21 + smod.22 + smod.23 + smod.30, 
        data = train[[i]],
        mtry = tuned_hyper[[i]][1,1], 
        min.node.size = tuned_hyper[[i]][1,2], 
        replace = tuned_hyper[[i]][1,3], 
        num.trees = tuned_hyper[[i]][1,5],
        respect.unordered.factors = "order",
        probability = TRUE,
        seed = 23
    )

    tuned_hyper[[i]]$rmse.final.prob <- sqrt(rf3[[i]]$prediction.error)

    saveRDS(rf3[[i]], file = paste("Outputs/", countries[[i]], "_10km_school_count_RF_Pred.rds", sep = ""))

}

p4 <- list()
predicted.probs <- list()
prob.pred <- list()
prob.pred.r <- list()
for (i in 1:length(countries)){
    p4[[i]] <- predict(rf3[[i]], country.data[[i]])
    predicted.probs[[i]] <- cbind(p4[[i]]$predictions, country.data[[i]])
    names(predicted.probs[[i]])[c(1,2)] <- c("prob_0", "prob_1")

    pred.sf <- st_as_sf(predicted.probs[[i]], coords = c("x","y"), crs = "ESRI:54009")
    st_crs(pred.sf) <- "ESRI:54009"
    pred.country <- st_intersection(pred.sf, AOI)
    pred.data <- cbind(st_drop_geometry(pred.country), st_coordinates(pred.country))
    
    pred.prob.r <- rast(predicted.probs[[i]][,c(4,5,1,2,3,6:21)], type = 'xyz', crs = "ESRI:54009")
    pred.prob.data <- as.data.frame(pred.prob.r, cell = TRUE, xy = TRUE)
    
    pred.prob.data$pred.schools <- ifelse(pred.prob.data$SCHOOLS == 1 | is.na(pred.prob.data$SCHOOLS), NA, pred.prob.data$prob_1)

    prob.r <- rast(pred.prob.data[,c(2,3,1,4:23)], type = 'xyz', crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs", digits = 6, extent = NULL)


    quant <- quantile(pred.prob.data[,"prob_1"], probs = c(0.8, 0.9, 0.95, 0.975, 0.99, 0.999, 1), na.rm = TRUE)

    pred.prob.data$sig.prob_0.2 <- ifelse(pred.prob.data$prob_1 < quant[1], NA, pred.prob.data$prob_1)
    pred.prob.data$sig.prob_0.1 <- ifelse(pred.prob.data$prob_1 < quant[2], NA, pred.prob.data$prob_1)
    pred.prob.data$sig.prob_0.05 <- ifelse(pred.prob.data$prob_1 < quant[3], NA, pred.prob.data$prob_1)
    pred.prob.data$sig.prob_0.025 <- ifelse(pred.prob.data$prob_1 < quant[4], NA, pred.prob.data$prob_1)
    pred.prob.data$sig.prob_0.01 <- ifelse(pred.prob.data$prob_1 < quant[5], NA, pred.prob.data$prob_1)
    pred.prob.data$sig.prob_0.001 <- ifelse(pred.prob.data$prob_1 < quant[6], NA, pred.prob.data$prob_1)

    prob.pred[[i]] <- pred.prob.data
    save(pred.prob.data, file = paste("Outputs/", countries[[i]], "_10km_school_count_RF_Pred_Prob.RData", sep = ""))

    prob.r <- rast(pred.prob.data[,c(2,3,1,4:29)], type = 'xyz', crs = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs", digits = 6, extent = NULL)
    prob.pred.r[[i]] <- prob.r
}

plot(prob.pred.r[[1]])

```



Any of the cropping and mapping stuff?