---
title: "schools_data_cleaning"
author: "Abi Riley"
date: "2023-09-18"
output: github_output
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-Up

```{r packages}

#install.packages(c("sf", "ggplot2", "stars", "ggpubr", "readr", "dplyr", "tidygeocoder", "tidyr"), dep = TRUE)

#install.packages("Rcpp")

## Load packages

library(ggplot2)

library(ggpubr)
#library(mapview)
library(readr)
library(dplyr)
library(tidygeocoder)
library(tidyr)
#library(sp)
library(Rcpp)
library(lwgeom)

library(remotes)
#install_github("r-spatial/sf", configure.args = "--with-proj-lib=$(brew --prefix)/lib/")

library(sf)
library(stars)

```

Was very difficult to install sf and stars, used this:

sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable
sudo apt-get update
sudo apt-get install libudunits2-dev libgdal-dev libgeos-dev libproj-dev libsqlite0-dev


# Schools Data

- schools_brgov: n = 216559 
- schools_unicef: n = 175704

```{r data, echo = FALSE}

schools_gov <- st_read("/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/schools_brgov_54009.gpkg")
schools_unicef <- st_read("/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/schools_unicef_54009.gpkg")

duplicates <- schools_gov[duplicated(schools_gov[,c("Longitude", "Latitude")]), ]

unique_gov <- schools_gov[!duplicated(schools_gov[,c("Escola", "Latitude", "Longitude")]), ]
full_gov <- subset(schools_gov, !is.na(Latitude) & !is.na(Longitude))
dups_gov <- full_gov[duplicated(full_gov[,c("Escola", "Latitude", "Longitude")]), ]
nrow(unique(dups_gov))
gov_dups_schools <- subset(schools_gov, Latitude %in% dups_gov$Latitude & Longitude %in% dups_gov$Longitude & Escola %in% dups_gov$Escola)
nrow(unique(gov_dups_schools))

missing_gov <- subset(schools_gov, is.na(Latitude) | is.na(Longitude))
non_dup_gov <- full_gov[!duplicated(full_gov[,c("Escola", "Latitude", "Longitude")]), ]
gov_schools <- rbind(missing_gov, non_dup_gov)


schools_unicef <- schools_unicef[!duplicated(schools_unicef[,c("name", "lat", "lon")]), ]
dups_unicef <- schools_unicef[duplicated(schools_unicef[,c("name", "lat", "lon")]), ]


```

# Missing Coordinates

Some government school entries do not have coordinate locations (n = 57817), instead we can geocode the addresses, which we have all of.

```{r missing-coords}

gov_missing <- subset(schools_gov, is.na(Latitude) | is.na(Longitude))
gov_full <- subset(schools_gov, !is.na(Latitude) & !is.na(Longitude))

names(gov_missing)
gov_missing_w_add2 <- subset(gov_missing, !is.na(Endereço))

```

Using a loop to batch run the "geocode" function, using the 'arcgis' method. 

WARNING: this takes AGES

```{r geocode}


school_missing <- subset(gov_missing_w_add[57001:57817,])
filled_missing <- geocode(gov_missing_w_add, Endereço, method = 'arcgis', lat = Latitude.new , long = Longitude.new)
save(filled_missing, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/filled_gov_schools_arcgis.RData")


for (i in 1:57){
    start <- floor((i-1)*1000) + 1
    end <- floor(i*1000)
    school_missing <- subset(gov_missing_w_add[start:end,])
    filled <- geocode(school_missing, Endereço, method = 'arcgis', lat = Latitude.new , long = Longitude.new)
    filled_missing <- rbind(filled_missing, filled2)

    save(filled_missing, file = paste("/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/filled_gov_schools_arcgis_", i, ".RData", sep = ""))
}


save(filled_missing, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/filled_gov_schools_arcgis.RData")

#load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/filled_gov_schools_arcgis.RData")

nrow(filled_missing)
nrow(unique(filled_missing[,c("Latitude.new", "Longitude.new")]))

```

Try this with the 2nd geocode method

```{r geocode1b}
names(filled_missing)

filled_missing$address <- paste(filled_missing$Escola, ", ", filled_missing$Município , ", ", filled_missing$UF, ", Brazil", sep = "")
#filled_missing$address

dups_filled <- filled_missing[duplicated(filled_missing[,c("Escola", "Latitude.new", "Longitude.new")]), ]
gov_dups_filled <- subset(filled_missing, Latitude.new %in% dups_filled$Latitude.new & Longitude.new %in% dups_filled$Longitude.new & Escola %in% dups_filled$Escola)
gov_unique_filled <- subset(filled_missing, !(Latitude.new %in% dups_filled$Latitude.new & Longitude.new %in% dups_filled$Longitude.new & Escola %in% dups_filled$Escola))


geocoded_duplicate <- geocode(gov_dups_filled, address,
                method = 'arcgis', lat = Latitude.new , long = Longitude.new)

geocoded_duplicate <- geocoded_duplicate[,c(1:20,24,25,23)]
names(geocoded_duplicate)[c(21,22)] <- c("Latitude.new", "Longitude.new")

nrow(geocoded_duplicate)
nrow(unique(geocoded_duplicate[,c("Latitude.new", "Longitude.new")]))
nrow(unique(geocoded_duplicate[,c("Escola", "Latitude.new", "Longitude.new")]))

dups_filled2 <- geocoded_duplicate[duplicated(geocoded_duplicate[,c("Escola", "Latitude.new", "Longitude.new")]), ]
gov_dups_filled2 <- subset(geocoded_duplicate, Latitude.new %in% dups_filled2$Latitude.new & Longitude.new %in% dups_filled2$Longitude.new & Escola %in% dups_filled2$Escola)
non_dup_gov2 <- geocoded_duplicate[!duplicated(geocoded_duplicate[,c("Escola", "Latitude", "Longitude")]), ]

geocoded_gov_schools <- rbind(gov_unique_filled, non_dup_gov2)


save(geocoded_gov_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/geocoded_filled2.RData")
load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/geocoded_filled2.RData")



```

Add geocoded indicator

```{r geocode-indicator}
filled_missing <- geocoded_gov_schools

names(filled_missing)
names(gov_full)

missing <- filled_missing[,-c(18,19)]
names(missing)
names(missing)[c(19,20)] <- c("Latitude", "Longitude")
missing <- missing[,c(1:17, 19,20,18)]

names(gov_full) == names(missing)

missing$geocoded <- "Yes"
gov_full$geocoded <- "No"

gov_schools <- rbind(missing, gov_full)

save(gov_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/all_coords_gov_schools_arcgis.RData")
#load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/all_coords_gov_schools_arcgis.RData")

nrow(gov_schools)
nrow(unique(gov_schools[,c("Escola", "Longitude", "Latitude")]))

gov_schools <- gov_schools[!duplicated(gov_schools[,c("Escola", "Latitude", "Longitude")]), ]

nrow(unique(non_dup_gov3[,c("Escola", "Longitude", "Latitude")]))

save(gov_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/all_coords_gov_schools_arcgis.RData")

ggplot() +
  stat_density_2d(data = gov_schools, 
                  mapping = aes(x = Longitude,
                                y = Latitude,
                                fill = stat(density)))

ggplot() + 
  geom_point(data = gov_schools, mapping = aes(x = Longitude, y = Latitude)) +
  stat_density_2d()


still_missing <- subset(gov_schools, is.na(Latitude) | is.na(Longitude))

```


# Centroid Schools

Some UNICEF school entries do not have the correct coordinates and are instead based on the centroid of the municipality.

We want to change this, with this priority order:

1. If available: Use government coords
2. Geocode the address
3. Leave as centroid, with an indicator

First, get centroid locations from each municipality

```{r centroid_locations}

Brazil <- st_read("/home/azureuser/cloudfiles/code/Users/ariley/Data/Shapefiles/Brazil_Admin_0_to_2/bra_admbnda_adm2_ibge_2020.shp")
st_crs(Brazil) <- "+proj=longlat +datum=WGS84 +no_defs +type=crs"
Brazil <- st_transform(Brazil, crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs")

sf_use_s2(FALSE)
muni_centroids <- st_centroid(Brazil, of_largest_polygon = TRUE)

centroids <- cbind(st_drop_geometry(muni_centroids), st_coordinates(muni_centroids))

```

Get municipality of each school.

```{r get_municipality_unicef}

schools_unicef$unicef_id <- 1:nrow(schools_unicef)

schools <- st_as_sf(schools_unicef, coords = c("lon", "lat"), crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs")
Brazil <- st_transform(Brazil, crs = st_crs(schools))
muni <- Brazil[,c("ADM2_PT")]
muni <- cbind(muni, 1:nrow(muni))
names(muni)[2] <- "ID.muni"
muni.df <- st_drop_geometry(muni)

muni.r <- st_rasterize(muni)
schools <- cbind(schools, st_extract(muni.r, schools)[1])
names(schools)[32] <- "ID.muni"

schools <- merge(schools, muni.df, by = "ID.muni")
names(schools)[c(33,34)] <- c("Municipality", "geom")

save(schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/unicef_schools_municipality.RData")
#load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/unicef_schools_municipality.RData")


schools <- cbind(st_drop_geometry(schools), st_coordinates(schools$geom))
schools.sf <- st_as_sf(schools)
schools.sf <- st_transform(schools.sf, crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs")
schools <- cbind(st_drop_geometry(schools.sf), st_coordinates(schools.sf))
schools <- schools[,-c(34,35)]

schools <- merge(schools, centroids, by.x = "Municipality", by.y = "ADM2_PT", all.x = TRUE, all.y = FALSE)

```

Measure distance to municipality centroid

```{r centroid_distance}

schools$x.diff <- abs(schools$X.x - schools$X.y)
schools$y.diff <- abs(schools$Y.x - schools$Y.y)

hist(schools$x.diff)
hist(schools$y.diff)

```
Get schools that are close to the centroids.

Clear drop off around 0.0005 degrees, approximately 55.5m.

n = 12653

```{r centroid_schools}
centroid_schools <- subset(schools, x.diff < 0.0005 & y.diff < 0.0005)
hist(centroid_schools$x.diff)
hist(centroid_schools$y.diff)

schools$centroid <- ifelse(schools$x.diff < 0.0005 & schools$y.diff < 0.0005, 1, 0)

names(schools)[c(34,35,43,44)] <- c("X", "Y", "X_centroid", "Y_centroid")

save(schools,  file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/unicef_schools_centroids.RData")
```

# Merge Datasets

Merge datasets on name and municipality. We can see that the locations are all very close, which is great.

```{r merge_schools}

load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/unicef_schools_centroids.RData")

schools.df <- st_drop_geometry(schools)

schools.df2 <- st_drop_geometry(gov_schools)

all_schools <- merge(schools.df, schools.df2, by.x = c("name", "Municipality"), by.y = c("Escola","Município"), all = TRUE)

par(mfrow = c(1,2))
plot(all_schools$lat, all_schools$Latitude)
plot(all_schools$lon, all_schools$Longitude)

save(all_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/all_merged_schools.RData")

load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/all_merged_schools.RData")

```

Now look at column names: what is the same? what is different? which has priority?

Exactly the same:
escola - name
Localização - school_region
dependência administrativa - school_type

Same meaning, different value:
latitude - lat
longitude - lon
porte da escola	- student_count

```{r data_columns}

names(all_schools)

```

# Choosing coordinates

Priority list:
1. Government locations (raw or geocoded)
2. UNICEF actual location
3. Geocoded UNICEF school
4. Centroid


```{r priority_coords}

all_schools$gov <- ifelse(is.na(all_schools$Latitude), "No", "Yes")

all_schools$X.chosen <- ifelse(is.na(all_schools$Longitude), all_schools$lon, all_schools$Longitude)
all_schools$Y.chosen <- ifelse(is.na(all_schools$Latitude), all_schools$lat, all_schools$Latitude)


unique(all_schools[,c("gov", "geocoded", "centroid")])

all_schools$geocoded <- all_schools$geocoded %>% replace_na("No")
all_schools$centroid <- all_schools$centroid %>% replace_na(0)

unique(all_schools[,c("gov", "geocoded", "centroid")])


```

Geocode school locations that are currently centroids, n = 2837

Make address and then use geocode

```{r geocode_centroid_schools}

centroid_schools <- subset(all_schools, gov == "No" & geocoded == "No" & centroid == 1)

centroid_schools$address <- paste(centroid_schools$name, ", ", centroid_schools$Municipality, ", ", centroid_schools$ADM1_PT, ", Brazil", sep = "")
centroid_schools$address

geocoded_centroid_schools <- geocode(centroid_schools, address,
                method = 'arcgis', lat = Latitude.new , long = Longitude.new)

geocoded_centroid_schools$Latitude.new
geocoded_centroid_schools$Longitude.new

save(geocoded_centroid_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/geocoded_centroids.RData")
load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/geocoded_centroids.RData")


geocoded_centroid_schools$X.chosen <- geocoded_centroid_schools$Longitude.new
geocoded_centroid_schools$Y.chosen <- geocoded_centroid_schools$Latitude.new

geocoded_centroid_schools$geocoded <- "Yes_2nd"
geocoded_centroid_schools <- geocoded_centroid_schools[,names(all_schools)]

other_schools <- subset(all_schools, !(gov == "No" & geocoded == "No" & centroid == 1))
all_schools <- rbind(other_schools, geocoded_centroid_schools)


unique(is.na(all_schools$X.chosen))
unique(is.na(all_schools$Y.chosen))


```

Final, all locations: N = 304,972

Label each one with where it came from 

```{r coord_labels}
unique(all_schools$geocoded)

all_schools$coord_source <- ifelse(all_schools$geocoded == "Yes_2nd", "2nd_geocode", 
    ifelse(all_schools$gov == "Yes" & all_schools$geocoded == "No", "gov", 
        ifelse(all_schools$gov == "No", "unicef", "1st_geocode")))

unique(all_schools$coord_source)

save(all_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/complete_all_schools.RData")

load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/complete_all_schools.RData")

nrow(all_schools)
nrow(unique(all_schools[,c("name", "X.chosen", "Y.chosen")]))


geo1 <- subset(all_schools, geocoded == "Yes_2nd")
nrow(geo1)
nrow(unique(geo1[,c("X.chosen", "Y.chosen")]))

unique(all_schools$geocoded)

names(all_schools)
dups <- all_schools[duplicated(all_schools[,c("name", "X.chosen", "Y.chosen")]), ]
dup_all <- subset(all_schools, X.chosen %in% dups$X.chosen & Y.chosen %in% dups$Y.chosen & name %in% dups$name)

non_dup_all <- all_schools[!duplicated(all_schools[,c("name", "X.chosen", "Y.chosen")]), ]
nrow(non_dup_all)

all_schools <- non_dup_all
save(all_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/complete_all_schools2.RData")

nrow(all_schools)

#st_write(all_schools, "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/complete_all_schools2.csv")

```

## Issue with 2 points
 2 geocoded points are outside of Brazil, both from the first geocoding.

 We will instead geocode these points using the school name and Municipality

```{r geocode3}
load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/complete_all_schools2.RData")


all_schools$address <- paste(all_schools$name, ", ", all_schools$Municipality, ", ", all_schools$ADM1_PT, ", Brazil", sep = "")

wrong_schools <- subset(all_schools, Endereço %in% c("AVENIDA EDESIO DE CARVALHO, 200 VILA MARCELA. 56350-000 Petrolina - PE.", "CAIXA POSTAL COMUNITARIA AVENIDA NORTE, 01 - LOTEAMENTO BEIJA-FLOR - CHA DE CRUZ, 21 LOTEAMENTO BEIJAFLOR. CHA DE CRUZ. 53590-990 Abreu e Lima - PE."))
right_schools <- subset(all_schools, !(Endereço %in% c("AVENIDA EDESIO DE CARVALHO, 200 VILA MARCELA. 56350-000 Petrolina - PE.", "CAIXA POSTAL COMUNITARIA AVENIDA NORTE, 01 - LOTEAMENTO BEIJA-FLOR - CHA DE CRUZ, 21 LOTEAMENTO BEIJAFLOR. CHA DE CRUZ. 53590-990 Abreu e Lima - PE.")))

wrong_schools[,c("X.chosen","Y.chosen")] <- geocode(wrong_schools, address, method = 'arcgis', lat = Latitude.new , long = Longitude.new)[,c("Longitude.new", "Latitude.new")]

all_schools <- rbind(right_schools, wrong_schools)

save(all_schools, file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/correct_all_schools.RData")

load(file = "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/correct_all_schools.RData")

nrow(all_schools)
nrow(unique(all_schools[,c("X.chosen", "Y.chosen")]))


nrow(all_schools)
nrow(unique(all_schools[,c("Latitude")]))

duplicates <- all_schools[duplicated(all_schools[,c("name","X.chosen","Y.chosen")]), ]
plot(duplicates$Longitude, duplicates$Latitude)


```


# Plot

```{r plot}

Brazil <- st_read("/home/azureuser/cloudfiles/code/Users/ariley/Data/Shapefiles/BR_UF_2021/BR_UF_2021.shp")
st_crs(Brazil)

Brazil <- st_transform(Brazil, crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs")

unique(is.na(all_schools$X.chosen))

hist(all_schools$X.chosen)
hist(all_schools$Y.chosen)

unique(is.na(all_schools$X.chosen))

ggplot(Brazil) +
  geom_sf() + coord_sf() +
  geom_point(all_schools, mapping = aes(x = X.chosen, y = Y.chosen, color = coord_source))

names(all_schools)

schools_sf <- st_as_sf(all_schools, coords = c("X.chosen", "Y.chosen"), crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs")

st_write(schools_sf, "/home/azureuser/cloudfiles/code/Users/ariley/Data/Schools/Other_Schools/complete_all_schools2.csv")


```